Estimating Predictive Rate-Distortion via Neural Variational Inference

### 2019-03-08 paper.tex
revising

Predicting the future from past observations is a key problem across science.
Constructing models that predict future observations is fundamental method of making and testing scientific discoveries.
Understanding and predicting dynamics has been a fundamental goal of physics for centuries.
In engineering, devices often have to predict the future of their environment to perform efficiently.
Living organisms need to make inferences about their environment and its future for survival.
Prediction permeats the behavior of organisms at all levels:
\textbf{TODO talk more about bacteria, animals, language, ....}

@@

%regularization: from finite data, want low-complexity model
%available data is finite, can only infer approximation to real dynamic
%In many cases, all we have is sample observation trajectories.

Optimal Causal Estimation (OCE, \citet{still-optimal-2010}) addresses the problem of estimating Predictive Rate Distortion from a finite sample of observation trajectories.
It does so by constructing a matrix of observed frequencies of different pairs of past and future observations.
However, this method faces a series of challenges~\citep{marzen-predictive-2016}.
One is the curse of dimensionality:
Modeling long dependencies requires storing an exponential number of observations, which quickly becomes intractable for current computation methods.
This exponential growth is particularly problematic when dealing with processes with a large state space.
For instance, the number of words in a human language as found in large-scale text data easily exceeds $10^5$, making storing and manipulating counts of longer word sequences very challenging.
A second challenge is that of overfitting:
%As OCE relies on counts of full past and future observation sequences,
When deploying a predictive model constructed via OCE on new data to predict upcoming observations, such a model can only succeed when the past sequences occurred in the sample that OCE was applied to.
This is because OCE relies on counts of full past and future observation sequences; it has no means to generalize to unseen past sequences.

@@

%\paragraph{Marginal $q$}
%For $q$, we choose the family of \emph{Neural Autoregressive Flows} \citep{}. 
%This is a parametric family of distributions that allows efficient estimation of the probability density and its gradients.
%This method widely generalizes a family of prior methods \citep{, kingma-improving-2016, papamakarios-masked-2017}, offering efficient estimation while surpassing prior methods in expressivity.

%TODO here is some more stuff:
%There are several motivations for studying this predictive rate-distortion tradeoff.
%One is that the shape of the tradeoff can provide insights into the structure of the process (TODO citations).
%%(CITE) show that phase transitions in the tradeoff curve reveal ...
%Second, this problem formalizes tradeoffs observed in real-world situations, where the goal is to predict the next observations in a sequence.
%Statistically modeling natural language, constructing models of the probability distribution over different sentences in a language, is often implemented as the problem of sequentially predicting the next word from the prior context.
%There is a tradeoff between the computational cost of representing the prior context, and accurately modeling the conditional distribution over the next word.
%Third, studying tis tradeoff may provide simpler but faithful models of processes (TODO citations).
%
%

@@

The problem of predictive rate-distortion has been formalized by a range of studies.
A principled and general formalization is provided by applying the Information Bottleneck idea \citep{still-optimal-2010, marzen-predictive-2016}:
We will write $\overleftarrow{X}$ for the past $X_{\leq 0}$, and $\overrightarrow{X}$ for the future $X_{> 0}$, following \cite{marzen-predictive-2016}.
We consider random variables $Z$, called \emph{codes}, that summarise the past and are used by the observer to predict the future.
Formally, $Z$ needs to be independent pf the future $\overrightarrow{X}$ conditional on the past $\overleftarrow{X}$. Symbolically, 
\begin{equation}
	Z \bot \overrightarrow{X} | \overleftarrow{X}
\end{equation}
This is equivalent to the requirement that $Z - \overleftarrow{X} - \overrightarrow{X}$ be a Markov chain.
This formalizes the idea that the code is computed by an observer from the past, without having access to the future.

The \emph{rate} of the code $Z$ is the mutual information between $Z$ and the past: $I[Z, \overleftarrow{X}]$.
By the Channel Coding Theorem, this describes the channel capacity that the observer requires in order to transform past observations into the code $Z$.

The \emph{distortion} is the loss in predictive accuracy when predicting from $Z$, relative to optimal prediction from the full past $\overleftarrow{X}$.
In the Information Bottleneck formalization, this
%the distortion measure is computed based on the entropy:
%\begin{equation}
%H[\overrightarrow{X}|Z] - H[\overrightarrow{X}|\overleftarrow{X}]
%\end{equation}
%This
is equivalent to the amount of mutual information between past and future that is not captured by $Z$:
\begin{equation}
	I[\overleftarrow{X}, \overrightarrow{X}|Z]
\end{equation}
Due to the Markov condition, the distortion measure satisfies the relation
\begin{equation}
	I[\overleftarrow{X}, \overrightarrow{X}|Z] = I[\overleftarrow{X}, \overrightarrow{X}] - I[\overrightarrow{X}, Z]
\end{equation}
For a fixed process $(X_t)_t$, minimizing the distortion is equivalent to maximizing the mutual information between the code and the future $I[\overrightarrow{X}, Z]$. We will refer to $I[\overrightarrow{X}, Z]$ as the \emph{predictiveness} of the code $Z$.

The rate-distortion tradeoff then chooses $Z$ to %minimize rate at bounded distortion:
%$$\min_{Z : I[\overleftarrow{X}, \overrightarrow{X}|Z] \leq d} I[Z, \overleftarrow{X}]$$
%, or -- equivalently --
 minimize distortion at bounded rate:
\begin{equation}
\min_{Z :  I[Z, \overleftarrow{X}] \leq d}  I[\overleftarrow{X}, \overrightarrow{X}|Z] 
\end{equation}
or -- equivalently -- maximize predictiveness at bounded rate:
\begin{equation}
\max_{Z :  I[Z, \overleftarrow{X}] \leq d}  I[Z, \overrightarrow{X}] 
\end{equation}
Equivalently, for each  $\beta \geq 0$, we study the problem 
\begin{equation}\label{eq:ib}
	\max_{Z} \left( I[\overrightarrow{X}, Z] - \beta \cdot I[Z, \overleftarrow{X}] \right)
\end{equation}
where the scope of the maximization is the class of all random variables $Z$ such that $Z \rightarrow \overleftarrow{X} \rightarrow \overrightarrow{X}$ is a Markov chain. %, that is, $Z$ is independent of the future given the past.

The objective (\ref{eq:ib}) is equivalent to the Information Bottleneck \citep{tishby-information-1999}, applied to the past and future of a stochastic process.

When $\beta \geq 1$, then (\ref{eq:ib}) is always optimized by a trivial $Z$ with zero rate and zero predictiveness, which follows from the Data Processing Inequality.
When $\beta = 0$, any lossless code optimizes the problem.
Therefore, we will be concerned with the situation where $\beta \in (0,1)$.

%
%
%Predictive rate-distortion describes the tradeoff between the cost of representing features of the past, and the quality of predictions in the future.
%Let $(X_t)$ be a stationary process.
%We consider feature representations $Z$ of the past $\overleftarrow{X}$ that are used to predict the future $\overrightarrow{X}$.
%There is a tradeoff between the complexity of $Z$ and the success of predicting the future:
%More complex representations of the past can reduce future prediction error -- up to a certain limit given by the entropy rate of the process. %This limit corresponds to perfect modeling, and results in Crutchfield and colleagues' notion of complexity.
%
%The complexity of the representation $Z$, or the \emph{rate}, is formalized as the mutual information between the past and the representation $Z$: $I[Z, \overleftarrow{X}]$.
%Different distortion measures could be used as measures of prediction error.
%CITE maximize the entropy, or, equivalently, minimize the mutual information, between $Z$ and the future: $I[Z, \overleftarrow{X}]$.
%
%Thus, for any process $(X_t)$, we obtain a rate-distortion tradeoff curve.
%The achievable region is populated by the random variables $Z$ such that $Z \rightarrow \left(\overleftarrow{X}\right) \rightarrow \left(\overrightarrow{X}\right)$ is a Markov chain.
%The points on the tradeoff curve are obtained by solving, for each $\beta \in [0,1]$, the problem
%\begin{equation}\label{eq:ib}
%	\max_{Z} \left( I[\overrightarrow{X}, Z] - \beta \cdot I[Z, (\overleftarrow{X})] \right)
%\end{equation}
%where the scope of the maximization is the class of all random variables $Z$ such that $Z - \left(\overleftarrow{X}\right) - \left(\overrightarrow{X}\right)$ is a Markov chain, that is, $Z$ is independent of the future given the past.
%This is exactly the Information Bottleneck \cite{tishby-information-1999}, applied to the past and future of a stochastic process.
%
%

@@

The problem of estimating statistical complexity, and the $\epsilon$-machine, is related to Predictive Rate-Distortion.
\citet{marzen-predictive-2016} show that predictive rate-distortion is equivalent to the problem of compressing causal states, i.e. equivalence classes of pasts, to predict causal states of the backwards process, i.e., equivalence classes of futures.
Further, \cite{still-optimal-2010} show that, in the limit of $\beta \rightarrow 0$, the code $Z$ that optimizes Predictive Rate-Distortion turns into the causal states.

@@

The main prior method for estimating predictive rate-distortion from data is Optimal Causal Estimation (OCE, \citet{still-optimal-2010}).
This method approximates predictive rate-distortion using two approximations:
First, it replaces semi-infinite pasts and futures with bounded-length contexts $(X_{-M, -1}), (X_{0, M-1})$:
\begin{equation}\label{eq:ib-oce}
	\max_{Z} \left( I[X_{-M\dots -1}, Z] - \beta \cdot I[Z, X_{0\dots M-1}] \right)
\end{equation}
Second, it estimates information measures directly from the observed counts of $(X_{-M, -1}), (X_{0, M-1})$ using the plug-in estimator of mutual information.
With such an estimator, the problem in (\ref{eq:ib}) can be solved using a variant of the Blahut-Arimoto algorithm \citep{tishby-information-1999}, obtaining an encoder $P(Z|X_{-M...-1})$ that maps each observed past sequence $X_{-M...-1}$ to a distribution over a (finite) set of codes $Z$. From this, one obtains a corresponding decoder $P(X_{0...M}|Z)$ via Bayes' Rule.

Two main challenges have been noted in prior work:
First, solving the problem for a finite empirical sample leads to overfitting, overestimating the amount of structure in the process.
\citet{still-optimal-2010} address this by subtracting an asymptotic correction term that becomes valid in the limit of large $M$ and $\beta \rightarrow 0$, when the codebook $P(Z|\overleftarrow{X})$ becomes deterministic, and which allows them to select a deterministic codebook of an appropriate complexity.
This leaves open how to obtain estimates outside of this regime, when the codebook can be far from deterministic.

@@

A further challenge is that OCE is infeasible if the number of required codewords is too large, again because OCE requires constructing a matrix whose rows and columns are indexed by the codewords and observed sequences.
Given that storing and manipulating matrices greater than $10^5 \times 10^5$ is currently not feasible, a setting where $I[Z, \overleftarrow{X}] > log(10^5) \approx 11.5$ cannot be captured with OCE.

@@

\begin{proposition} 
The KL Divergence between $P(Z|X)$ and $q(Z)$, averaged over $X$, upper-bounds the mutual information between $Z$ and the past observations:
\begin{equation}\label{ineq2}
\begin{aligned}
	\mathbb{E}_{X}\left[ \operatorname{D_{KL}}(\infdiv{P(Z|X)}{q(Z)})\right] &=   \mathbb{E}_{X} \mathbb{E}_{Z|X} \log \frac{P_\phi(Z|X_{-M\dots -1})}{q(Z)}  \\
	& \geq  \mathbb{E}_{X} \mathbb{E}_{Z|X} \log \frac{P(Z|X_{-M\dots -1})}{P(Z)}  \\
	& = I[X_{-M\dots 0},Z].
\end{aligned}
\end{equation}
Equality is attained if and only if $q(Z)$ is equal to the true marginal $P(Z) = \E_{X_{-M \dots -1}} \phi(Z|X_{-M \dots -1})$.
\end{proposition}

Therefore, the predictive rate-distortion objective is equivalent to
\begin{equation}\label{eq:bound}
	\min_{\phi, \psi, q}	\left[-	\mathbb{E}_{z \sim \phi(X_{-M\dots-1})}\left[\log P(X_{0\dots M} | z)\right] + \beta \cdot \operatorname{D_{KL}}(\infdiv{\phi(X_{-M\dots-1})}{q})\right],
\end{equation}
where $\phi, \psi, q$ range over all triples of the appropriate types described above.
To carry out this optimization, we will restrict to a powerful family of parametric families of function approximators, within which we will optimize the objective with gradient descent.
The values obtained for the resulting $\phi, \psi, q$ satisfy the bounds in (\ref{ineq1}-\ref{ineq2}), and---if the family of approximators is sufficiently rich---can come close to turning these into equalities.

@@

Following standard practice in machine learning, we split the data into a training and a held-out set, using the training set for optimizing the parameters.
After every pass through the training set, the objective is evaluated on the held-out set; optimization terminates once the value on the held-out set does not improve any more.

\paragraph{Evaluating}
After optimizing the parameters on a set of observed trajectories, we evaluate it on a held-out set of trajectories to estimate (\ref{eq:bound}).
Given parameters for $\phi, \psi, q$, we evaluate the objective~(\ref{eq:bound}) on the dataset by taking, for each timeseries $X_{-M}...X_M$, a single sample of $z  \sim \mathcal{N}(\mu, \sigma^2)$ and computing
\begin{equation}\label{eq:bound-mc}
	\frac{1}{N}	\sum_{X_{-M...M} \in TestData}	\log P(X_{0\dots M} | z_{-M...-1}) + \beta \cdot \log \frac{\mathcal{N}(\mu, \sigma^2)(z)}{q(z)}
\end{equation}
When evaluating (\ref{eq:bound}) on unseen data this way, the resulting value is guaranteed to be an \emph{upper} bound on the true predictive rate-distortion due to (\ref{ineq1}-\ref{ineq2}), up to sampling error introduced by sampling $z$ and the finite size of the held-out dataset.
It is important to note that this sampling error is different from the overfitting issue affecting OCE: 
(\ref{eq:bound-mc}) is an \emph{unbiased} estimator of an upper bound.
The magnitude of the estimation error can be estimated using standard methods such as bootstrapping, cross-validation, and taking more samples of $z$---unlike overfitting affecting OCE, which \emph{biases} the values obtained.

@@

As a computationally more efficient method, we defined a range of plausible hyperparameters based both on experience reported in the literature, and considerations of computational efficiency.
These parameters are shown in Table~\ref{tab:nprd-hyperparameters}.
We then randomly sampled, for each of the processes that we experimented on, combinations of $\beta$ and these hyperparameters to run NPRD on.

@@

\begin{table}
\begin{tabular}{l||lll}
	& Section~\ref{sec:tractable} &   POS & Words\\ \hline\hline
	Embedding Dimension & 50, 100 &       &  150      \\
	LSTM Dimension & 32           &       &    512     \\ \hline
%	LSTM Layers & 1 \\ \hline
	Dropout rate & 0.0, 0.1, 0.2  &       &      0.4     \\
	Input Dropout & 0  &       &        0.2             \\ \hline
	Adam Learning Rate & 0.0001, 0.0005, 0.001 (4 times), 0.002, 0.004  &       &       0.00005, 0.0001, 0.0005, 0.001     \\
	Batch Size & 16, 32, 64  &       &       32          \\ \hline
	Flow Length & 1, 2  &       &          4              \\
	Flow Type & DSF, DDSF  &       &       DSF, DDSF             \\
	Flow Dimension & 32, 64, 128, 512  &       &    ???        \\
	Flow Layers & 2  &       &                 ?????          \\
\end{tabular}
	\caption{NPRD Hyperparameters}\label{tab:nprd-hyperparameters}
\end{table}

@@

We applied NPRD to these processes by training on 3M random trajectories of length 30, and using 3000 additional trajectories as held-out data.
For each process, we ran NPRD 1000 times for randomly choices of $\beta \in [0,1]$.
Due to computational constraints, when running OCE, we limited sample size to 3000 trajectories for estimation and as held-out data.
Following \cite{marzen-predictive-2016}, we ran OCE for $M=1,...,5$.

@@

\paragraph{Recovering Causal States}
Does NPRD lead to inpretable codes $Z$?
To answer this, we further investigated the NPRD approximation to the Random Insertion Process (RIP), obtained in the previous paragraph.
The $\epsilon$-machine was computed by  \cite{marzen-predictive-2016} and is given in Figure~\ref{fig:rip-machine} (right).
The process has three caual states:
A represents those pasts where the future starts with $1^{2k}0$ ($k = 0, 1, 2, \dots$) -- these are the pasts ending in either $001$ or $10111^m$ ($m= 0, 1, 2, \dots$). % but cannot go in with 10
B represents those pasts ending in 10 -- the future has to start with 01 or 11.
C represents those pasts ending in either 00 or 01 -- the future has to start with $1^{2k+1}0$ ($k = 0, 1, 2, \dots$).
%These states represent the 
%As itThis $\epsilon$-machine is the smallest unifilar

The analytical solution to the predictive rate-distortion problem was computed by \cite{marzen-predictive-2016}.
At $\beta > 0.5$, the optimal solution collapses $A$ and $B$ into a single codeword, while all three states are mapped to separate codewords for $\beta \leq 0.5$.

Does NPRD recover this picture?
We applied PCA to samples from $Z$ computed at two different values of $\beta$, $\beta = 0.25$ and $\beta = 0.6$, corresponding to the two points marked in the rate-distortion figure.
The first two principal components are shown in Figure~\ref{fig:latent}.
Samples are colored by the causal states corresponding to the pasts of the trajectories that were encoded into the respective points by NPRD.
On the left, obtained at $\beta=0.6$, the states A and B are collapsed, as expected.
On the right, obtained at $\beta=0.25$, the three causal states are reflected as distinct modes of $Z$.
Note that, at finite $M$, a fraction of pasts is ambiguous between the green and blue causal states; these are colored in black and NPRD maps them into a region between the modes corresponding to these states.

@@

Russian

>>>>

Predicting the future from past observations is a key problem across science.
Constructing models that predict future observations is fundamental method of making and testing scientific discoveries.
Understanding and predicting dynamics has been a fundamental goal of physics for centuries.
In engineering, devices often have to predict the future of their environment to perform efficiently.
Living organisms need to make inferences about their environment and its future for survival.
%Prediction has been argued to permeat the behavior of organisms at all levels:
%\textbf{TODO some more about biological systems, language, ...?}

@@

Optimal Causal Estimation (OCE, \citet{still-optimal-2010}) addresses the problem of estimating Predictive Rate Distortion from a finite sample of observation trajectories.
It does so by constructing a matrix of observed frequencies of different pairs of past and future observations.
However, this method faces a series of challenges~\citep{marzen-predictive-2016}.
One is the curse of dimensionality:
Modeling long dependencies requires storing an exponential number of observations, which quickly becomes intractable for current computation methods.
This exponential growth is particularly problematic when dealing with processes with a large state space.
For instance, the number of words in a human language as found in large-scale text data easily exceeds $10^5$, making storing and manipulating counts of longer word sequences very challenging.
A second challenge is that of overfitting:
When deploying a predictive model constructed via OCE on new data to predict upcoming observations, such a model can only succeed when the past sequences occurred in the sample that OCE was applied to.
This is because OCE relies on counts of full past and future observation sequences; it has no means to generalize to unseen past sequences.

@@

%\paragraph{Marginal $q$}
%For $q$, we choose the family of \emph{Neural Autoregressive Flows} \citep{}. 
%This is a parametric family of distributions that allows efficient estimation of the probability density and its gradients.
%This method widely generalizes a family of prior methods \citep{, kingma-improving-2016, papamakarios-masked-2017}, offering efficient estimation while surpassing prior methods in expressivity.
%TODO here is some more stuff:
%There are several motivations for studying this predictive rate-distortion tradeoff.
%One is that the shape of the tradeoff can provide insights into the structure of the process (TODO citations).
%%(CITE) show that phase transitions in the tradeoff curve reveal ...
%Second, this problem formalizes tradeoffs observed in real-world situations, where the goal is to predict the next observations in a sequence.
%Statistically modeling natural language, constructing models of the probability distribution over different sentences in a language, is often implemented as the problem of sequentially predicting the next word from the prior context.
%There is a tradeoff between the computational cost of representing the prior context, and accurately modeling the conditional distribution over the next word.
%Third, studying tis tradeoff may provide simpler but faithful models of processes (TODO citations).

@@

The problem of predictive rate-distortion has been formalized by a range of studies.
A principled and general formalization is provided by applying the Information Bottleneck idea \citep{tishby-information-1999, still-optimal-2010, marzen-predictive-2016}:
We will write $\overleftarrow{X}$ for the past $X_{\leq 0}$, and $\overrightarrow{X}$ for the future $X_{> 0}$, following \cite{marzen-predictive-2016}.
We consider random variables $Z$, called \emph{codes}, that summarise the past and are used by the observer to predict the future.
Formally, $Z$ needs to be independent pf the future $\overrightarrow{X}$ conditional on the past $\overleftarrow{X}$. Symbolically, 
\begin{equation}\label{eq:markov}
	Z \bot \overrightarrow{X} | \overleftarrow{X}
\end{equation}
This is equivalent to the requirement that $Z - \overleftarrow{X} - \overrightarrow{X}$ be a Markov chain.
This formalizes the idea that the code is computed by an observer from the past, without having access to the future.

The \emph{rate} of the code $Z$ is the mutual information between $Z$ and the past: $\operatorname{I}[Z, \overleftarrow{X}]$.
By the Channel Coding Theorem, this describes the channel capacity that the observer requires in order to transform past observations into the code $Z$.

The \emph{distortion} is the loss in predictive accuracy when predicting from $Z$, relative to optimal prediction from the full past $\overleftarrow{X}$.
In the Information Bottleneck formalization, this
is equivalent to the amount of mutual information between past and future that is not captured by $Z$:
\begin{equation}
	\operatorname{I}[\overleftarrow{X}, \overrightarrow{X}|Z]
\end{equation}
Due to the Markov condition, the distortion measure satisfies the relation
\begin{equation}
	\operatorname{I}[\overleftarrow{X}, \overrightarrow{X}|Z] = \operatorname{I}[\overleftarrow{X}, \overrightarrow{X}] - \operatorname{I}[\overrightarrow{X}, Z]
\end{equation}
For a fixed process $(X_t)_t$, minimizing the distortion is equivalent to maximizing the mutual information between the code and the future $\operatorname{I}[\overrightarrow{X}, Z]$. We will refer to $\operatorname{I}[\overrightarrow{X}, Z]$ as the \emph{predictiveness} of the code $Z$.

The rate-distortion tradeoff then chooses $Z$ to %minimize rate at bounded distortion:
%$$\min_{Z : \operatorname{I}[\overleftarrow{X}, \overrightarrow{X}|Z] \leq d} \operatorname{I}[Z, \overleftarrow{X}]$$
%, or -- equivalently --
 minimize distortion at bounded rate:
\begin{equation}
\min_{Z :  \operatorname{I}[Z, \overleftarrow{X}] \leq d}  \operatorname{I}[\overleftarrow{X}, \overrightarrow{X}|Z] 
\end{equation}
or -- equivalently -- maximize predictiveness at bounded rate:
\begin{equation}
\max_{Z :  \operatorname{I}[Z, \overleftarrow{X}] \leq d}  \operatorname{I}[Z, \overrightarrow{X}] 
\end{equation}
Equivalently, for each  $\lambda \geq 0$, we study the problem 
\begin{equation}\label{eq:ib}
	\max_{Z} \left( \operatorname{I}[\overrightarrow{X}, Z] - \lambda \cdot \operatorname{I}[Z, \overleftarrow{X}] \right)
\end{equation}
where the scope of the maximization is the class of all random variables $Z$ such that $Z \rightarrow \overleftarrow{X} \rightarrow \overrightarrow{X}$ is a Markov chain. %, that is, $Z$ is independent of the future given the past.

The objective (\ref{eq:ib}) is equivalent to the Information Bottleneck \citep{tishby-information-1999}, applied to the past and future of a stochastic process.
The coefficient $\lambda$ indicates how strongly a high rate $\operatorname{I}[Z, \overleftarrow{X}]$ is penalized; higher values of $\lambda$ result in lower rates and thus lower values of predictiveness.

The largest achievable predictiveness $\operatorname{I}[\overrightarrow{X}, Z]$ is equal to $\operatorname{I}[\overrightarrow{X}, \overleftarrow{X}]$, which is known as the \emph{excess entropy} of the process~\citep{feldman-synchronizing-2004}.
Due to the Markov condition~(\ref{eq:markov}) and the Data Processing Inequality, predictiveness of a code $Z$ is always upper-bounded by the rate:
\begin{equation}
\operatorname{I}[\overrightarrow{X}, Z] \leq \operatorname{I}[Z, \overleftarrow{X}] 
\end{equation}
As a consequence, when $\lambda \geq 1$, then (\ref{eq:ib}) is always optimized by a trivial $Z$ with zero rate and zero predictiveness.
When $\lambda = 0$, any lossless code optimizes the problem.
Therefore, we will be concerned with the situation where $\lambda \in (0,1)$.

%
%
%Predictive rate-distortion describes the tradeoff between the cost of representing features of the past, and the quality of predictions in the future.
%Let $(X_t)$ be a stationary process.
%We consider feature representations $Z$ of the past $\overleftarrow{X}$ that are used to predict the future $\overrightarrow{X}$.
%There is a tradeoff between the complexity of $Z$ and the success of predicting the future:
%More complex representations of the past can reduce future prediction error -- up to a certain limit given by the entropy rate of the process. %This limit corresponds to perfect modeling, and results in Crutchfield and colleagues' notion of complexity.
%
%The complexity of the representation $Z$, or the \emph{rate}, is formalized as the mutual information between the past and the representation $Z$: $\operatorname{I}[Z, \overleftarrow{X}]$.
%Different distortion measures could be used as measures of prediction error.
%CITE maximize the entropy, or, equivalently, minimize the mutual information, between $Z$ and the future: $\operatorname{I}[Z, \overleftarrow{X}]$.
%
%Thus, for any process $(X_t)$, we obtain a rate-distortion tradeoff curve.
%The achievable region is populated by the random variables $Z$ such that $Z \rightarrow \left(\overleftarrow{X}\right) \rightarrow \left(\overrightarrow{X}\right)$ is a Markov chain.
%The points on the tradeoff curve are obtained by solving, for each $\beta \in [0,1]$, the problem
%\begin{equation}\label{eq:ib}
%	\max_{Z} \left( \operatorname{I}[\overrightarrow{X}, Z] - \beta \cdot \operatorname{I}[Z, (\overleftarrow{X})] \right)
%\end{equation}
%where the scope of the maximization is the class of all random variables $Z$ such that $Z - \left(\overleftarrow{X}\right) - \left(\overrightarrow{X}\right)$ is a Markov chain, that is, $Z$ is independent of the future given the past.
%This is exactly the Information Bottleneck \cite{tishby-information-1999}, applied to the past and future of a stochastic process.
%
%

@@

The problem of estimating statistical complexity, and the $\epsilon$-machine, is related to Predictive Rate-Distortion.
\citet{marzen-predictive-2016} show that predictive rate-distortion is equivalent to the problem of compressing causal states, i.e. equivalence classes of pasts, to predict causal states of the backwards process, i.e., equivalence classes of futures.
Further, \cite{still-optimal-2010} show that, in the limit of $\lambda \rightarrow 0$, the code $Z$ that optimizes Predictive Rate-Distortion turns into the causal states.

@@

The main prior method for estimating predictive rate-distortion from data is Optimal Causal Estimation (OCE, \citet{still-optimal-2010}).
This method approximates predictive rate-distortion using two approximations:
First, it replaces semi-infinite pasts and futures with bounded-length contexts $(X_{-M, -1}), (X_{0, M-1})$:
\begin{equation}\label{eq:ib-oce}
	\max_{Z} \left( \operatorname{I}[X_{-M\dots -1}, Z] - \lambda \cdot \operatorname{I}[Z, X_{0\dots M-1}] \right)
\end{equation}
Second, it estimates information measures directly from the observed counts of $(X_{-M, -1}), (X_{0, M-1})$ using the plug-in estimator of mutual information.
With such an estimator, the problem in (\ref{eq:ib}) can be solved using a variant of the Blahut-Arimoto algorithm \citep{tishby-information-1999}, obtaining an encoder $P(Z|X_{-M...-1})$ that maps each observed past sequence $X_{-M...-1}$ to a distribution over a (finite) set of codes $Z$. From this, one obtains a corresponding decoder $P(X_{0...M}|Z)$ via Bayes' Rule.

Two main challenges have been noted in prior work:
First, solving the problem for a finite empirical sample leads to overfitting, overestimating the amount of structure in the process.
\citet{still-optimal-2010} address this by subtracting an asymptotic correction term that becomes valid in the limit of large $M$ and $\lambda \rightarrow 0$, when the codebook $P(Z|\overleftarrow{X})$ becomes deterministic, and which allows them to select a deterministic codebook of an appropriate complexity.
This leaves open how to obtain estimates outside of this regime, when the codebook can be far from deterministic.

@@

A further challenge is that OCE is infeasible if the number of required codewords is too large, again because OCE requires constructing a matrix whose rows and columns are indexed by the codewords and observed sequences.
Given that storing and manipulating matrices greater than $10^5 \times 10^5$ is currently not feasible, a setting where $\operatorname{I}[Z, \overleftarrow{X}] > log(10^5) \approx 11.5$ cannot be captured with OCE.

@@

\begin{proposition} 
The KL Divergence between $P(Z|X)$ and $q(Z)$, averaged over $X$, upper-bounds the mutual information between $Z$ and the past observations:
\begin{equation}\label{ineq2}
\begin{aligned}
	\mathbb{E}_{X}\left[ \operatorname{D_{KL}}(\infdiv{P(Z|X)}{q(Z)})\right] &=   \mathbb{E}_{X} \mathbb{E}_{Z|X} \log \frac{P_\phi(Z|X_{-M\dots -1})}{q(Z)}  \\
	& \geq  \mathbb{E}_{X} \mathbb{E}_{Z|X} \log \frac{P(Z|X_{-M\dots -1})}{P(Z)}  \\
	& = \operatorname{I}[X_{-M\dots 0},Z].
\end{aligned}
\end{equation}
Equality is attained if and only if $q(Z)$ is equal to the true marginal $P(Z) = \E_{X_{-M \dots -1}} \phi(Z|X_{-M \dots -1})$.
\end{proposition}

Therefore, the predictive rate-distortion objective is equivalent to
\begin{equation}\label{eq:bound}
	\min_{\phi, \psi, q}	\left[-	\mathbb{E}_{z \sim \phi(X_{-M\dots-1})}\left[\log P(X_{0\dots M} | z)\right] + \lambda \cdot \operatorname{D_{KL}}(\infdiv{\phi(X_{-M\dots-1})}{q})\right],
\end{equation}
where $\phi, \psi, q$ range over all triples of the appropriate types described above.
To carry out this optimization, we will restrict to a powerful family of parametric families of function approximators, within which we will optimize the objective with gradient descent.
The values obtained for the resulting $\phi, \psi, q$ satisfy the bounds in (\ref{ineq1}-\ref{ineq2}), and---if the family of approximators is sufficiently rich---can come close to turning these into equalities.

@@

Following standard practice in machine learning, we split the data into three partitions (\emph{training set}, \emph{validation set}, and \emph{test set}).
We use the training set for optimizing the parameters.
After every pass through the training set, the objective is evaluated on the validation set; optimization terminates once the value on the held-out set does not improve any more.
After this, we use the test set to estimate rate and predictiveness of the fully optimized model (see next paragraph).

\paragraph{Evaluating}
After optimizing the parameters on a set of observed trajectories, we evaluate it on the test set to estimate (\ref{eq:bound}).
Given parameters for $\phi, \psi, q$, we evaluate the objective~(\ref{eq:bound}) on the dataset by taking, for each timeseries $X_{-M}...X_M$, a single sample of $z  \sim \mathcal{N}(\mu, \sigma^2)$ and computing
\begin{equation}\label{eq:bound-mc}
	\frac{1}{N}	\sum_{X_{-M...M} \in TestData}	\log P(X_{0\dots M} | z_{-M...-1}) + \lambda \cdot \log \frac{\mathcal{N}(\mu, \sigma^2)(z)}{q(z)}
\end{equation}
When evaluating (\ref{eq:bound}) on unseen data this way, the resulting value is guaranteed to be an \emph{upper} bound on the true predictive rate-distortion due to (\ref{ineq1}-\ref{ineq2}), up to sampling error introduced by sampling $z$ and the finite size of the held-out dataset.
It is important to note that this sampling error is different from the overfitting issue affecting OCE: 
(\ref{eq:bound-mc}) is an \emph{unbiased} estimator of an upper bound.
The magnitude of the estimation error can be estimated using standard methods such as bootstrapping, cross-validation, and taking more samples of $z$---unlike overfitting affecting OCE, which \emph{biases} the values obtained.

@@

As a computationally more efficient method, we defined a range of plausible hyperparameters based both on experience reported in the literature, and considerations of computational efficiency.
These parameters are shown in Table~\ref{tab:nprd-hyperparameters}.
We then randomly sampled, for each of the processes that we experimented on, combinations of $\lambda$ and these hyperparameters to run NPRD on.

@@

\begin{table}
\begin{tabular}{l||lll}
	& Section~\ref{sec:tractable} &   POS & Words\\ \hline\hline
	Embedding Dimension & 50, 100 &       &  150      \\
	LSTM Dimension & 32           &       &    512     \\ \hline
%	LSTM Layers & 1 \\ \hline
	Dropout rate & 0.0, 0.1, 0.2  &       &      0.4     \\
	Input Dropout & 0  &       &        0.2             \\ \hline
	Adam Learning Rate & 0.0001, 0.0005, 0.001 (4 times), 0.002, 0.004  &       &       0.00005, 0.0001, 0.0005, 0.001     \\
	Batch Size & 16, 32, 64  &       &       32          \\ \hline
	Flow Length & 1, 2  &       &          4              \\
	Flow Type & DSF, DDSF  &       &       DSF, DDSF             \\
	Flow Dimension & 32, 64, 128, 512  &       &    TODO        \\
	Flow Layers & 2  &       &                 TODO         \\
\end{tabular}
	\caption{NPRD Hyperparameters}\label{tab:nprd-hyperparameters}
\end{table}

@@

We applied NPRD to these processes by training on 3M random trajectories of length 30, and using 3000 additional trajectories for validation and test data.
For each process, we ran NPRD 1000 times for randomly choices of $\lambda \in [0,1]$.
Due to computational constraints, when running OCE, we limited sample size to 3000 trajectories for estimation and as held-out data.
Following \cite{marzen-predictive-2016}, we ran OCE for $M=1,...,5$.

@@

\paragraph{Recovering Causal States}
Does NPRD lead to inpretable codes $Z$?
To answer this, we further investigated the NPRD approximation to the Random Insertion Process (RIP), obtained in the previous paragraph.
The $\epsilon$-machine was computed by  \cite{marzen-predictive-2016} and is given in Figure~\ref{fig:rip-machine} (right).
The process has three caual states:
A represents those pasts where the future starts with $1^{2k}0$ ($k = 0, 1, 2, \dots$) -- these are the pasts ending in either $001$ or $10111^m$ ($m= 0, 1, 2, \dots$). % but cannot go in with 10
B represents those pasts ending in 10 -- the future has to start with 01 or 11.
C represents those pasts ending in either 00 or 01 -- the future has to start with $1^{2k+1}0$ ($k = 0, 1, 2, \dots$).

The analytical solution to the predictive rate-distortion problem was computed by \cite{marzen-predictive-2016}.
At $\lambda > 0.5$, the optimal solution collapses $A$ and $B$ into a single codeword, while all three states are mapped to separate codewords for $\lambda \leq 0.5$.

Does NPRD recover this picture?
We applied PCA to samples from $Z$ computed at two different values of $\lambda$, $\lambda = 0.25$ and $\lambda = 0.6$, corresponding to the two points marked in the rate-distortion figure.
The first two principal components are shown in Figure~\ref{fig:latent}.
Samples are colored by the causal states corresponding to the pasts of the trajectories that were encoded into the respective points by NPRD.
On the left, obtained at $\lambda=0.6$, the states A and B are collapsed, as expected.
On the right, obtained at $\lambda=0.25$, the three causal states are reflected as distinct modes of $Z$.
Note that, at finite $M$, a fraction of pasts is ambiguous between the green and blue causal states; these are colored in black and NPRD maps them into a region between the modes corresponding to these states.

@@

		Russian {\tiny ($\alpha=0.5$, $\beta=2.3$, $E_0=1.7$)}
